Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/background.tex
fixed usage of '32 byte' without a hyphen
fixed the captions for the graphs, and a few of the graphs themselves
fixed some missing hyphens in cache types

When a cache contains the data being sought, then this is called a cache
\n{hit}. When it does not, it is called a cache \n{miss}, and the next largest
cache in the hierarchy is queried for the data. There are three types of cache
miss: \n{compulsory} or \n{cold-start} misses, \n{conflict} misses and
\n{capacity} misses. A compulsory miss occurs when the data is first fetched
from memory. Since it has not been accessed before, it cannot already be in the
cache. A conflict miss occurs when the addressing scheme used by the cache
causes two separate memory locations to be mapped to the same area of the cache.
This only occurs in direct mapped or set-associative cache, described below.
Finally, a capacity miss occurs when the data sought has been ejected from the
cache due to a lack of space.

Caches range from direct mapped to fully associative. In a direct mapped cache,
every memory address has an exact position in the cache where it can be found.
Since the cache is smaller than main memory, many addresses map to the same position.
Fully associative is the opposite of this: data can be stored anywhere in the
cache. A set-associative cache is a mixture of both: more than one piece of data
can be stored in the same place. A 4-way set-associative cache allows four items
to be stored in the same place; when a fifth item needs to be stored in that
position, one of the items is ejected. The item may be ejected randomly, in a
specific order, or the \n{least recently used} item may be removed.

When data is loaded into the cache, data near it is loaded with it. This set of
data is stored in a \n{cache line} or \n{cache block}. The associativity and
length of the cache line dictate the mapping of memory addresses to cache
storage. Figure \vref{Addressing} shows how a memory address is divided.

\begin{figure}[h]
\centering
\includegraphics{cache_addressing}
\caption{Addressing}
\label{Addressing}
\end{figure}

This cache is a 2MB direct mapped cache.  There is a 32 byte cache line, which
is indexed by the 5 bit \n{offset}. There are $2^{16}$ cache lines, which are
indexed by the middle 16 bits of the address. The most significant 11 bits form
the \n{tag}. A cache line is tagged to check if the data stored at an index is
the same as the data requested. For a fully associative cache, there is no
index, and all the bits except the offset make up the tag. A fully associative
cache has no conflict misses for this reason; similarly, a direct mapped cache
has no capacity misses, since even when full, misses are due to conflicts.

>>>>

When a cache contains the data being sought, then this is called a cache
\n{hit}. When it does not, it is called a cache \n{miss}, and the next largest
cache in the hierarchy is queried for the data. There are three types of cache
miss: \n{compulsory} or \n{cold-start} misses, \n{conflict} misses and
\n{capacity} misses. A compulsory miss occurs when the data is first fetched
from memory. Since it has not been accessed before, it cannot already be in the
cache. A conflict miss occurs when the addressing scheme used by the cache
causes two separate memory locations to be mapped to the same area of the cache.
This only occurs in direct-mapped or set-associative cache, described below.
Finally, a capacity miss occurs when the data sought has been ejected from the
cache due to a lack of space.

Caches range from direct-mapped to fully-associative. In a direct-mapped cache,
every memory address has an exact position in the cache where it can be found.
Since the cache is smaller than main memory, many addresses map to the same position.
Fully-associative is the opposite of this: data can be stored anywhere in the
cache. A set-associative cache is a mixture of both: more than one piece of data
can be stored in the same place. A 4-way set-associative cache allows four items
to be stored in the same place; when a fifth item needs to be stored in that
position, one of the items is ejected. The item may be ejected randomly, in a
specific order, or the \n{least recently used} item may be removed.

When data is loaded into the cache, data near it is loaded with it. This set of
data is stored in a \n{cache line} or \n{cache block}. The associativity and
length of the cache line dictate the mapping of memory addresses to cache
storage. Figure \vref{Address Division} shows how a memory address is divided.

\begin{figure}[h]
\centering
\includegraphics{cache_addressing}
\caption{Address Division}
\label{Address Division}
\end{figure}

This cache is a 2MB direct-mapped cache.  There is a 32-byte cache line, which
is indexed by the 5 bit \n{offset}. There are $2^{16}$ cache lines, which are
indexed by the middle 16 bits of the address. The most significant 11 bits form
the \n{tag}. A cache line is tagged to check if the data stored at an index is
the same as the data requested. For a fully-associative cache, there is no
index, and all the bits except the offset make up the tag. A fully-associative
cache has no conflict misses for this reason; similarly, a direct-mapped cache
has no capacity misses, since even when full, misses are due to conflicts.

### 2005-08-14 docs/tech_report/heap.tex
fixed some missing hyphens in cache types

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, showing base heapsort's poor temporal
locality, with the 8-heap having the fewest misses. This continues into the
level 2 cache, where the flattest slope belongs to the 8-heap. The difference
between fully associative and direct mapped is only minor, with the fully
associative performing slightly better than the direct mapped cache.

@@

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the memory-tuned heapsort's do in
the results presented here. LaMarca's results indicate that the memory-tuned heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimized algorithm would not suffer as much.

>>>>

The level 1 cache results, shown in Figure \ref{heap level 1 misses}, show the
difference between the two types of heap. The slope of the base heapsort result
is much greater than all the others, showing base heapsort's poor temporal
locality, with the 8-heap having the fewest misses. This continues into the
level 2 cache, where the flattest slope belongs to the 8-heap. The difference
between fully-associative and direct-mapped is only minor, with the
fully-associative performing slightly better than the direct-mapped cache.

@@

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the memory-tuned heapsort's do in
the results presented here. LaMarca's results indicate that the memory-tuned heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct-mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimized algorithm would not suffer as much.

### 2005-08-14 docs/tech_report/merge.tex
fixed some missing hyphens in cache types

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully associative caches should not have these misses, and the results should be
correspondingly lower. The double mergesorts should have
better performance than their counterparts in level 1 misses, though there
should be a slight increase in level 2 misses. It is expected that they should
perform slightly faster as a result.

@@

\plot{merge}{}{}{Results shown are for a direct mapped cache.}{Results shown are
for a direct mapped cache.}{}{}

@@

\label{mergesort replacement policy}
It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully associative cache will perform slightly better is false. In actual fact,
the direct mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully associative cache removes the least recently used keys from the cache
to make room, while the direct mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

>>>>

The number of memory references is also $O(NlogN)$, and mergesort's cache
performance should be very regular, so long as the data set fits in the cache. Once
the data set exceeds the cache, the number of cache misses should leap. For
multi-mergesort, they should flatten out after this, perhaps with a slight
slope to take into account the conflict misses of the aligned array segments. The
fully-associative caches should not have these misses, and the results should be
correspondingly lower. The double mergesorts should have
better performance than their counterparts in level 1 misses, though there
should be a slight increase in level 2 misses. It is expected that they should
perform slightly faster as a result.

@@

\plot{merge}{}{}{Results shown are for a direct-mapped cache.}{Results shown are
for a direct-mapped cache.}{}{}

@@

\label{mergesort replacement policy}
It is notable that our prediction of half a miss per key for multi-mergesort,
which was taken from LaMarca, is exactly right. However, our assertion that the
fully-associative cache will perform slightly better is false. In actual fact,
the direct-mapped cache performs significantly better (see Figure
\vref{multi-mergesort level 2 misses}). This is due to the replacement policy:
our fully-associative cache removes the least recently used keys from the cache
to make room, while the direct-mapped caches are designed to make best use of
the keys while they are in the cache. They have, effectively, an optimised
replacement policy.

### 2005-08-14 docs/tech_report/quick.tex
fixed some missing hyphens in cache types

\plot{quick}{}{}{Results shown are for a direct mapped cache.}{Results shown are
for a direct mapped cache.}{}{}{}

@@

The results show again (see Section \ref{mergesort replacement policy}) that
direct mapped caches can perform better than fully associative ones. This is
true in the case of all the quicksorts, as can be seen in Figures \ref{base
quicksort (no median) level 2 misses} to \ref{multi-quicksort (sequential
search) level 2 misses} graph that this is true in the case of all the
quicksorts. As in Section \ref{mergesort replacement policy}, this is due to the
replacement policy of the fully-associative cache.

>>>>

\plot{quick}{}{}{Results shown are for a direct-mapped cache.}{Results shown are
for a direct-mapped cache.}{}{}{}

@@

The results show again (see Section \ref{mergesort replacement policy}) that
direct-mapped caches can perform better than fully-associative ones. This is
true in the case of all the quicksorts, as can be seen in Figures \ref{base
quicksort (no median) level 2 misses} to \ref{multi-quicksort (sequential
search) level 2 misses} graph that this is true in the case of all the
quicksorts. As in Section \ref{mergesort replacement policy}, this is due to the
replacement policy of the fully-associative cache.
